---
title: "Introduction to Data Science Group Project Task 2"
author: "K9"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
# !!! NOTE THAT: by default, in the pdf file generated, it does not show the code
# chunk, if want to show chunk, change the line below to **echo = TRUE**
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(knitr)
library(purrr)
library(broom)
```

In this section, we will discuss the extent each continent working toward the goal
of reducing the proportion of youth not in employment, education or training (hereafter
referred to as NEET).

We would first draw a graph to demonstrate the change of NEET for each continent 
with the year using the line graph.

```{r import data, message=FALSE, warning=FALSE}
youth_neet <- read.csv("Exported_data/youth_neet_cleaned.csv")
gdp <- read.csv("Exported_data/gdp_per_capita_cleaned.csv")
```

```{r Continent NEET against Year (line chart)}
youth_neet %>%
  group_by(Continent, Year) %>%
  summarise(
    NEET_avg = weighted.mean(Percentage_NEET, Population, na.rm = T),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = Year, y = NEET_avg, colour = Continent)) +
  geom_point() +
  geom_line() +
  labs(
    x = "Year",
    y = "Percentage",
    title = "Comparison of NEET"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(
    size = 12,
    hjust = 0
  ))
```

\newpage

We then would like to see the regression line for the NEET.

```{r analysis}
# Calculate weighted averages by continent and year
# This accounts for population differences between countries
youth_summary <- youth_neet %>%
  group_by(Continent, Year) %>%
  summarise(NEET_avg = weighted.mean(Percentage_NEET, Population, na.rm = TRUE),
            .groups = "drop") %>%
  ungroup()

# Define color scheme for continents
continent_colors <- c(
  "Africa"        = "#F8766D",
  "Asia"          = "#B79F00",
  "Europe"        = "#00BA38",
  "North America" = "#00BFC4",
  "Oceania"       = "#619CFF",
  "South America" = "#F564E3"
)

# Calculate regression slopes
slope_neet_year <- youth_summary %>%
  group_by(Continent) %>%
  do({
    model <- lm(NEET_avg ~ Year, data = .)
    tidy_model <- tidy(model)
    slope_row <- tidy_model %>% filter(term == "Year")
    
    # Create result data frame with slope and p-value
    data.frame(
      Continent = unique(.$Continent),
      slope = slope_row$estimate,      # Rate of change per year
      p = slope_row$p.value,           # Statistical significance
      stringsAsFactors = FALSE
    )
  }) %>%
  ungroup()

# Create comprehensive comparison table
slope_neet_year <- slope_neet_year %>%
  mutate(
    # Binary significance at 5% level
    significance = ifelse(p < 0.05, "Yes", "No"),
    
    # Star notation for significance levels
    sig_label = case_when(
      p < 0.001 ~ "***",   # Highly significant
      p < 0.01 ~ "**",     # Very significant
      p < 0.05 ~ "*",      # Significant
      TRUE ~ "n.s."             # Not significant
    )
  )

# Display comparison table
# Shows side-by-side results for hypothesis testing
kable(
  slope_neet_year %>%
    select(
      Continent,
      `Slope` = slope,
      `P-value` = p,
      `Significance` = sig_label,
      `Significant at 5%? (with)` = significance
    ),
  digits = 4,
  caption = "Hypothesis Testing Results: With and Without Outliers (IQR multiplier: 1.0)"
)


# Prepare slope annotations for visualization
# Creates formatted labels showing slope and significance
slope_data <- slope_neet_year %>%
  select(Continent, slope, sig_label) %>%
  mutate(
    slope_label = paste0(
      "Slope: ", sprintf("%.3f", slope),  # Format to 3 decimal places
      " (", sig_label, ")"                # Add significance stars
    )
  )

# Create visualization
# Shows trends for all continents with regression lines
ggplot(youth_summary, aes(x = Year, y = NEET_avg)) +
  geom_point(aes(colour = Continent), alpha = 0.7, size = 3) +
  geom_smooth(aes(colour = Continent, fill = Continent),
              method = "lm", se = TRUE, linewidth = 1) +
  geom_text(data = slope_data,
            aes(label = slope_label, colour = Continent),
            x = -Inf, y = Inf, hjust = -0.1, vjust = 1.2,
            size = 3.5, fontface = "bold") +
  scale_colour_manual(values = continent_colors) +
  scale_fill_manual(values = continent_colors) +
  facet_wrap(~ Continent, scales = "free_y") +
  labs(
    title = "Youth Not in EET: Trend by Continent",
    subtitle = "Weighted averages with linear regression trend lines",
    x = "Year",
    y = "NEET (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14, face = "bold"),
    strip.text = element_text(size = 12, face = "bold")
  )

```

\newpage

Then we would like to see how education and employment each contribute to the EET, 
one would create a table to illustrate how they correlated

```{r Contribution to NEET, message=FALSE}
# First, we merge the school enrollment and youth unemployment data into the
# youth summary data frame. Ignore NAs so that the weighted average calculation
# would be more calculating from the rest of the data.
youth_summary <- youth_neet %>%
  group_by(Continent, Year) %>%
  summarise(
    SE_avg = weighted.mean(School_Enrollment, Population, na.rm = TRUE),
    Unemp_avg = weighted.mean(Youth_Unemployment, Population, na.rm = T),
    .groups = "keep"
  ) %>%
  right_join(., youth_summary) %>%
  ungroup()


# Then we would like to calculate the correlation between NEET rate and school
# enrollment rate and youth unemployment rate.

# I used AI for this part, specifically how to turn the cor.test() result into
# a data frame.

Cor_NEET_Con <- (function() {
  SE_corr <- youth_summary %>%
    group_by(Continent) %>%
    summarise(
      test = list(cor.test(NEET_avg, SE_avg, method = "spearman"))
    ) %>%
    mutate(
      Corr_SE = map_dbl(test, ~ round(.x$estimate, 4)),
      P_Value_SE = map_dbl(test, ~ round(.x$p.value, 4))
    ) %>%
    select(-test)

  Unemp_corr <- youth_summary %>%
    group_by(Continent) %>%
    summarise(
      test = list(cor.test(NEET_avg, Unemp_avg, method = "spearman"))
    ) %>%
    mutate(
      Corr_Unemp = map_dbl(test, ~ round(.x$estimate, 4)),
      P_Value_Unemp = map_dbl(test, ~ round(.x$p.value, 4))
    ) %>%
    select(-test)

  full_join(
    SE_corr, Unemp_corr,
    by = "Continent"
  )
})()

kable(Cor_NEET_Con,
  format = "markdown",
  caption = "Correlation of NEET with Youth Unemployment and School Enrollment (Continental)"
)
```

If we do not group the data by continent rather seeing them as a whole,then we 
get the following table

```{r, warning=FALSE}
Cor_NEET_World <- (function() {
  NEET_SE <- cor.test(youth_neet$Percentage_NEET,
    youth_neet$School_Enrollment,
    method = "spearman",
    use = "complete.obs"
  )

  NEET_Unemp <- cor.test(youth_neet$Percentage_NEET,
    youth_neet$Youth_Unemployment,
    method = "spearman",
    use = "complete.obs",
  )
  
  data.frame(
    variable = c("SE", "Unemployment"),
    correlation = round(c(NEET_Unemp$estimate, NEET_SE$estimate), 4),
    p_value = round(c(NEET_Unemp$p.value, NEET_SE$p.value), 4)
  )
})()

kable(Cor_NEET_World,
  format = "markdown",
  caption = "Correlation of NEET with Youth Unemployment and School Enrollment (World)"
)
```

```{r merge GDP and NEET}
gdp_neet <- inner_join(gdp, youth_neet, by = c("Entity", "Code", "Year", "Continent"))

# Clean and restructure the merged dataset
gdp_neet <- gdp_neet %>% 
  select(Continent, Entity, Year, Gdp_per_capita, Percentage_NEET)

# Convert Continent to a factor for grouping and visualization
gdp_neet$Continent <- as.factor(gdp_neet$Continent)
```

# Global Correlation
```{r global_correlation, warning=FALSE}
# Perform correlation test at the global (country) level
# Tests the linear relationship between GDP per capita and NEET percentage
# across all countries and years in the dataset
global_cor_s <- cor.test(
  gdp_neet$Gdp_per_capita,
  gdp_neet$Percentage_NEET,
  method = "spearman"
)

global_cor_p <- cor.test(
  gdp_neet$Gdp_per_capita,
  gdp_neet$Percentage_NEET,
  method = "pearson"
)
# Display correlation results (r value, p-value, confidence interval)
global_cor_p
global_cor_s
```

# Global Plot
```{r global_plot}
ggplot(gdp_neet, aes(Gdp_per_capita, Percentage_NEET, color = Continent)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_x_log10() +
  labs(
    title = "GDP per Capita vs NEET (%) — Global",
    x = "GDP per Capita (log scale)",
    y = "NEET (%)"
  ) +
  theme_minimal()
```

# Continent Means + Correlation
```{r continent_means}
# Aggregate data to the continent level by calculating mean values
continent_means <- gdp_neet %>%
  group_by(Continent) %>%
  summarise(
    Mean_GDP  = mean(Gdp_per_capita),
    Mean_NEET = mean(Percentage_NEET),
    .groups   = "drop"
  )

# Perform correlation test on continent-level aggregated data
# This tests whether continents with higher average GDP have lower average NEET
continent_cor_p <- cor.test(
  continent_means$Mean_GDP,
  continent_means$Mean_NEET,
  method = "pearson"
)

continent_cor_s <- cor.test(
  continent_means$Mean_GDP,
  continent_means$Mean_NEET,
  method = "spearman"
)
```

# Continent Plot
```{r continent_plot}
ggplot(continent_means, aes(Mean_GDP, Mean_NEET)) +
  geom_point(size = 4, aes(color = Continent)) +
  geom_text(aes(label = Continent), vjust = -1) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_x_log10() +
  labs(
    title = "Mean GDP vs Mean NEET — Continent Level",
    x = "Mean GDP per Capita (log scale)",
    y = "Mean NEET (%)"
  ) +
  theme_minimal()
```

```{r results_table}
# Take siginificance level at 0.05
ALPHA <- 0.05
# Create a summary table comparing the two levels of analysis
Correlation_GDP_NEET <- data.frame(
  Level = c("Global (Countries)", "Continent Means (6 points)"),
  Pearson_r = c(global_cor_p$estimate, continent_cor_p$estimate),
  Spearman_r = c(global_cor_s$estimate, continent_cor_s$estimate),
  
  # Extract p-values to assess statistical significance
  p_value = c(global_cor_p$p.value, continent_cor_p$p.value),
  N = c(nrow(gdp_neet), nrow(continent_means)),
  Significant = c(global_cor_p$p.value < ALPHA,
                  continent_cor_p$p.value < ALPHA)
)

# Display the results table for easy comparison
Correlation_GDP_NEET
```















