---
title: "Task 1"
author: "CJD"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Librarys:
```{r}
library(tidyverse)
library(ggplot2)
```


Set WD and import data
```{r}
gdp_per_capita_df <- read.csv("Exported_data/gdp_per_capita_cleaned.csv")
```


Calculate the total GDP of each continent
```{r}
continent_gdp_totals <- gdp_per_capita_df %>%
  group_by(Continent, Year) %>%
  summarise(
    Total_GDP = sum(Gdp, na.rm = TRUE)
  ) %>%
  ungroup()
```


Plot a graph of GDP for each continent, vs time
```{r}
continent_gdp_totals %>% ggplot(aes(x = Year, y = Total_GDP, colour = Continent)) +
  geom_point() +
  geom_line() +
  labs(title = "GDP/Year for 6 continents", y = "GDP ($)")
```


I decided this graph could show more if they all started at an index of 100
```{r}
continent_gdp_totals <- continent_gdp_totals %>%
  group_by(Continent) %>%
  mutate(
    GDP_Index = Total_GDP / first(Total_GDP) * 100
  )
```



We now replot the graph, using normalised values
```{r}
continent_gdp_totals %>% ggplot(aes(x = Year, y = GDP_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_text(
  data = continent_gdp_totals %>% filter(Year == 2021),
  aes(label = round(GDP_Index, 1)),
  hjust = -0.3,
  size = 3,
  show.legend = FALSE) +
  expand_limits(x = 2022) +
  labs(title = "GDP Index/Year for 6 continents", y = "GDP Index")
```


Calculate the weighted and unweighted growth of each continent
```{r}
continents_growth_df <- gdp_per_capita_df %>%
  group_by(Continent, Year) %>%
  summarise(
    Weighted_Growth = sum(Growth * Gdp, na.rm = TRUE) / sum(Gdp, na.rm = TRUE),
    Avg_Growth = mean(Growth, na.rm = TRUE)
  ) %>%
  ungroup()
```


Plot the growth graphs of each continent
```{r}
continents_growth_df %>% ggplot(aes(x = Year, y = Weighted_Growth, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth (weighted)/Year for 6 continents", y = "GDP Growth (%)")

continents_growth_df %>% ggplot(aes(x = Year, y = Avg_Growth, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth (avg)/Year for 6 continents", y = "GDP Growth (%)")
```


I decided to do a similar thing with growth about index
```{r}
continents_growth_df <- continents_growth_df %>%
  group_by(Continent) %>%
  mutate(
    Avg_Growth_Index = Avg_Growth / first(Avg_Growth) * 100,
    Weighted_Growth_Index = Weighted_Growth / first(Weighted_Growth) * 100
  )
```


Plotting the graphs
```{r}
continents_growth_df %>% ggplot(aes(x = Year, y = Weighted_Growth_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth Index (weighted)/Year for 6 continents", y = "GDP Growth (%)")

continents_growth_df %>% ggplot(aes(x = Year, y = Avg_Growth_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth Index (avg)/Year for 6 continents", y = "GDP Growth (%)")
```




REF:

28/11/2025 - asked ChatGPT - "i have text near the last points on the x axis, how can i make it so it doesn't overlap the axis"













Plot ldc growth, each separate country in a  certain continent (6 graphs)
Create a ldc database

```{r}
ldc <- gdp_per_capita_df %>%
  filter (Ldc == TRUE)
```

Plot the graphs

```{r}
ggplot(ldc, aes(x = Year, y = Growth, color = Entity)) +
  geom_line() +
  facet_wrap(~ Continent) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme(legend.position = "none")

```

Do a weighted average of ldc growth rate for each contine, then plot a graph of each continent and how they compare

```{r}
ldc_weighted <- ldc %>%
  group_by(Year, Continent) %>%
  mutate(
    weighted_growth = sum(Gdp_per_capita * Growth, na.rm = TRUE) / sum(Gdp_per_capita, na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(ldc_weighted, aes(x = Year, y = weighted_growth, colour = Continent)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ Continent) +
  theme(legend.position = "none") +
  geom_hline(yintercept = 7, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") 


```

Identify which LDCs have had a growth that reached 7% and plot their growth

```{r}
ldc_7 <- ldc %>%
  group_by(Entity) %>%
  filter(any(Growth >= 7)) %>%
  ungroup()

ggplot (ldc_7, aes(x = Year, y = Growth, color = Entity)) + 
          geom_line() + 
          facet_wrap (~ Continent) +
          theme_minimal() + 
          labs(
            title = "GDP Growth of LDC that reached 7%",
            x = "Year",
            y = "GDP Growth(%)"
          )
```


REF:

28/11/2025 - asked ChatGPT - "“I have a facet plot in ggplot with some continents having tons of lines and some only one. How can I adjust the y-axis scales so each graph uses the right space for its lines and they don’t look squished or empty?"









#Find the 3 countries with most GDP growth, and 3 countries with most GDP decline 

# Top 3 growth
top_growth <- gdp_per_capita_df %>%
  group_by(Entity) %>%
  summarise(Avg_Growth = mean(Growth, na.rm = TRUE)) %>%
  arrange(desc(Avg_Growth)) %>%
  slice(1:3)

# Top 3 decline
top_decline <- gdp_per_capita_df %>%
  group_by(Entity) %>%
  summarise(Avg_Growth = mean(Growth, na.rm = TRUE)) %>%
  arrange(Avg_Growth) %>%
  slice(1:3)

#find the countries with biggest gdp in each continent 
#plot their graphs & analyse their trends, 
#as these will most likely be the drivers behind any fluctuations, 
#especially in weighted avg 

#find the countries with biggest gdp in each continent 
top_3_per_continent <- gdp_per_capita_df %>%
  group_by(Continent, Entity) %>%
  summarise(Total_GDP = sum(Gdp, na.rm = TRUE)) %>%
  arrange(Continent, desc(Total_GDP)) %>%
  group_by(Continent) %>%
  slice(1:3)

#plot their graphs 
filtered_df <- gdp_per_capita_df %>%
  semi_join(top_3_per_continent, by = c("Continent", "Entity"))

ggplot(filtered_df, aes(x = Year, y = Gdp, colour = Entity)) +
  geom_line() +
  facet_wrap(~ Continent, scales = "free_y") +
  labs(title = "Top 3 Countries by GDP per Capita (per Continent)", y = "GDP per capita")
#free y scaling - allows the y-axis of a plot to vary independently for each row of facets

#ANALYSIS
#Africa: Nigeria, South Africa, Algeria
#Asia: China, India, Japan
#Europe: Germany, UK, France. Germany leading 
#North America: US, Mexico, Canada
#Oceania: Australia, New Zealand, Papa New Guinea
#South America: Brazil, Argentina, Colombia. Brazil drives volatility, steady increase 
#in all continents apart from Africa, there is one clear country that is driving GDP
#why this is important to analyse? these countries are the ones that drive continent averages
#the weighted averages are heavily influenced by large economics (eg US, China, Germany) 
#spikes/dips in major economies explain most fluctations in continent graphs. this can definitely be seen in europe in thr last few years for example








