---
title: "Task 1"
author: "CJD"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Librarys:
```{r}
library(tidyverse)
library(ggplot2)
```


Set WD and import data
```{r}
gdp_per_capita_df <- read.csv("Exported_data/gdp_per_capita_cleaned.csv")
gdp_per_capita_df <- gdp_per_capita_df %>%
  filter(Year >= 2015)
```


Calculate the total GDP of each continent
```{r}
continent_gdp_totals <- gdp_per_capita_df %>%
  group_by(Continent, Year) %>%
  summarise(
    Total_GDP = sum(Gdp, na.rm = TRUE)
  ) %>%
  ungroup()
```


Plot a graph of GDP for each continent, vs time
```{r}
continent_gdp_totals %>% ggplot(aes(x = Year, y = Total_GDP, colour = Continent)) +
  geom_point() +
  geom_line() +
  labs(title = "GDP/Year for 6 continents", y = "GDP ($)")
```


I decided this graph could show more if they all started at an index of 100
```{r}
continent_gdp_totals <- continent_gdp_totals %>%
  group_by(Continent) %>%
  mutate(
    GDP_Index = Total_GDP / first(Total_GDP) * 100
  )
```



We now replot the graph, using normalised values
```{r}
continent_gdp_totals %>% ggplot(aes(x = Year, y = GDP_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_text(
  data = continent_gdp_totals %>% filter(Year == 2021),
  aes(label = round(GDP_Index, 1)),
  hjust = -0.3,
  size = 3,
  show.legend = FALSE) +
  expand_limits(x = 2022) +
  labs(title = "GDP Index/Year for 6 continents", y = "GDP Index")
```


Calculate the weighted and unweighted growth of each continent
```{r}
continents_growth_df <- gdp_per_capita_df %>%
  group_by(Continent, Year) %>%
  summarise(
    Weighted_Growth = sum(Growth * Gdp, na.rm = TRUE) / sum(Gdp, na.rm = TRUE),
    Avg_Growth = mean(Growth, na.rm = TRUE)
  ) %>%
  ungroup()
```


Plot the growth graphs of each continent
```{r}
continents_growth_df %>% ggplot(aes(x = Year, y = Weighted_Growth, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth (weighted)/Year for 6 continents", y = "GDP Growth (%)")

continents_growth_df %>% ggplot(aes(x = Year, y = Avg_Growth, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth (avg)/Year for 6 continents", y = "GDP Growth (%)")
```


I decided to do a similar thing with growth about index (realised is redundant, don't use in analysis)
```{r}
continents_growth_df <- continents_growth_df %>%
  group_by(Continent) %>%
  mutate(
    Avg_Growth_Index = Avg_Growth / first(Avg_Growth) * 100,
    Weighted_Growth_Index = Weighted_Growth / first(Weighted_Growth) * 100
  )
```


Plotting the graphs
```{r}
continents_growth_df %>% ggplot(aes(x = Year, y = Weighted_Growth_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 1) +
  labs(title = "GDP Growth Index (weighted)/Year for 6 continents", y = "GDP Growth (%)")

continents_growth_df %>% ggplot(aes(x = Year, y = Avg_Growth_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 1) +
  labs(title = "GDP Growth Index (avg)/Year for 6 continents", y = "GDP Growth (%)")
```




REF:

28/11/2025 - asked ChatGPT - "i have text near the last points on the x axis, how can i make it so it doesn't overlap the axis"














Create an LDC database

```{r}
ldc <- gdp_per_capita_df %>%
  filter (Ldc == TRUE)
```


Plot GDP growth by continent

```{r}
ggplot(ldc, aes(x = Year, y = Growth, color = Entity)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ Continent) +
  labs(
    title = "GDP Growth of LDCs by Continent",
    x = "Year",
    y = "GDP Growth (%)"
  )
```

Compute the weighted average GDP growth for each continent

```{r}
ldc_weighted <- ldc %>%
  group_by(Year, Continent) %>%
  summarise(
    weighted_growth =
      sum(Gdp_per_capita * Growth, na.rm = TRUE) /
      sum(Gdp_per_capita, na.rm = TRUE),
    .groups = "drop"
  )
```


Plot weighted GDP growth by continent with a 7% reference line

```{r}
ggplot(ldc_weighted, aes(x = Year, y = weighted_growth, color = Continent)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ Continent, scales = "free_y") +
  geom_hline(yintercept = 7, linetype = "dashed", color = "red") +
  labs(
    title = "Weighted GDP Growth of LDCs by Continent",
    x = "Year",
    y = "Weighted GDP Growth (%)"
  )
```


Identify which LDCs have had a growth that reached 7% and plot their growth

```{r}
ldc_7 <- ldc %>%
  group_by(Entity) %>%
  filter(any(Growth >= 7)) %>%
  ungroup()
```


Plot their graph

```{r}
ggplot (ldc_7, aes(x = Year, y = Growth, color = Entity)) + 
  geom_point() +
  geom_line() + 
  facet_wrap (~ Continent) +
  labs(
    title = "GDP Growth of LDCs that reached 7%",
    x = "Year",
    y = "GDP Growth(%)"
  )
```



Determine which countries have had the highest growth over the years

```{r}
ldc %>%
  group_by(Entity) %>%
  summarise(avg_growth = mean(Growth, na.rm = TRUE)) %>%
  arrange(desc(avg_growth)) 
```


REF:

28/11/2025 - asked ChatGPT - "“I have a facet plot in ggplot with some continents having tons of lines and some only one. How can I adjust the y-axis scales so each graph uses the right space for its lines and they don’t look squished or empty?"







Find the 3 countries with most GDP growth, and 3 countries with most GDP decline 

```{r}
# Top 3 growth
top_growth <- gdp_per_capita_df %>%
  group_by(Entity) %>%
  summarise(Avg_Growth = mean(Growth, na.rm = TRUE)) %>%
  arrange(desc(Avg_Growth)) %>%
  slice(1:3)

# Top 3 decline
top_decline <- gdp_per_capita_df %>%
  group_by(Entity) %>%
  summarise(Avg_Growth = mean(Growth, na.rm = TRUE)) %>%
  arrange(Avg_Growth) %>%
  slice(1:3)
```

Find the countries with biggest gdp in each continent 
```{r}
top_3_per_continent <- gdp_per_capita_df %>%
  group_by(Continent, Entity) %>%
  summarise(Total_GDP = sum(Gdp, na.rm = TRUE)) %>%
  arrange(Continent, desc(Total_GDP)) %>%
  group_by(Continent) %>%
  slice(1:3)
```

Plot their graphs 
```{r}
filtered_df <- gdp_per_capita_df %>%
  semi_join(top_3_per_continent, by = c("Continent", "Entity"))

ggplot(filtered_df, aes(x = Year, y = Gdp, colour = Entity)) +
  geom_line() +
  facet_wrap(~ Continent, scales = "free_y") +
  labs(title = "Top 3 Countries by GDP per Capita (per Continent)", y = "GDP per capita")
#free y scaling - allows the y-axis of a plot to vary independently for each row of facets
```




Graph to see if china and india are the drivers of asias growth
```{r}


top_2_per_continent <- gdp_per_capita_df %>%
  group_by(Continent, Entity) %>%
  summarise(Total_GDP = sum(Gdp, na.rm = TRUE)) %>%
  arrange(Continent, desc(Total_GDP)) %>%
  group_by(Continent) %>%
  slice(1:2) %>%
  ungroup()

get_plot_data <- function(x) {

  top_2_entities <- top_2_per_continent %>%
    filter(Continent == x) %>%
    pull(Entity)

  country_data <- gdp_per_capita_df %>%
    filter(Continent == x, Entity %in% top_2_entities)

  continent_avg <- continents_growth_df %>%
    filter(Continent == x) %>%
    rename(Growth = Avg_Growth) %>%
    mutate(Entity = "Continent Average")

  plot_data <- bind_rows(country_data, continent_avg)

  return(plot_data)
}

plot_drivers <- function(continent) {
  data <- get_plot_data(continent)

  ggplot(data, aes(x = Year, y = Growth, colour = Entity)) +
    geom_line() +
    geom_point() +
    labs(
      title = paste("Top 2 GDP Growth vs Avg -", continent),
      y = "Growth (%)"
    )
}

plot_drivers("Africa")
plot_drivers("Asia")
plot_drivers("Europe")
plot_drivers("North America")
plot_drivers("South America")
plot_drivers("Oceania")



```



ANALYSIS
Africa: Nigeria, South Africa, Algeria
Asia: China, India, Japan
Europe: Germany, UK, France. Germany leading 
North America: US, Mexico, Canada
Oceania: Australia, New Zealand, Papa New Guinea
South America: Brazil, Argentina, Colombia. Brazil drives volatility, steady increase 

in all continents apart from Africa, there is one clear country that is driving GDP
why this is important to analyse? these countries are the ones that drive continent verages
the weighted averages are heavily influenced by large economics (eg US, China, Germany) 
spikes/dips in major economies explain most fluctations in continent graphs. this can definitely be seen in europe in thr last few years for example
