---
title: "Task 1"
author: "CJD"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Librarys:
```{r}
library(tidyverse)
library(ggplot2)
```


Set WD and import data
```{r}
gdp_per_capita_df <- read.csv("Exported_data/gdp_per_capita_cleaned.csv")
```


Calculate the total GDP of each continent
```{r}
continent_gdp_totals <- gdp_per_capita_df %>%
  group_by(Continent, Year) %>%
  summarise(
    Total_GDP = sum(Gdp, na.rm = TRUE)
  ) %>%
  ungroup()
```


Plot a graph of GDP for each continent, vs time
```{r}
continent_gdp_totals %>% ggplot(aes(x = Year, y = Total_GDP, colour = Continent)) +
  geom_point() +
  geom_line() +
  labs(title = "GDP/Year for 6 continents", y = "GDP ($)")
```


I decided this graph could show more if they all started at an index of 100
```{r}
continent_gdp_totals <- continent_gdp_totals %>%
  group_by(Continent) %>%
  mutate(
    GDP_Index = Total_GDP / first(Total_GDP) * 100
  )
```



We now replot the graph, using normalised values
```{r}
continent_gdp_totals %>% ggplot(aes(x = Year, y = GDP_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_text(
  data = continent_gdp_totals %>% filter(Year == 2021),
  aes(label = round(GDP_Index, 1)),
  hjust = -0.3,
  size = 3,
  show.legend = FALSE) +
  expand_limits(x = 2022) +
  labs(title = "GDP Index/Year for 6 continents", y = "GDP Index")
```


Calculate the weighted and unweighted growth of each continent
```{r}
continents_growth_df <- gdp_per_capita_df %>%
  group_by(Continent, Year) %>%
  summarise(
    Weighted_Growth = sum(Growth * Gdp, na.rm = TRUE) / sum(Gdp, na.rm = TRUE),
    Avg_Growth = mean(Growth, na.rm = TRUE)
  ) %>%
  ungroup()
```


Plot the growth graphs of each continent
```{r}
continents_growth_df %>% ggplot(aes(x = Year, y = Weighted_Growth, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth (weighted)/Year for 6 continents", y = "GDP Growth (%)")

continents_growth_df %>% ggplot(aes(x = Year, y = Avg_Growth, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth (avg)/Year for 6 continents", y = "GDP Growth (%)")
```


I decided to do a similar thing with growth about index
```{r}
continents_growth_df <- continents_growth_df %>%
  group_by(Continent) %>%
  mutate(
    Avg_Growth_Index = Avg_Growth / first(Avg_Growth) * 100,
    Weighted_Growth_Index = Weighted_Growth / first(Weighted_Growth) * 100
  )
```


Plotting the graphs
```{r}
continents_growth_df %>% ggplot(aes(x = Year, y = Weighted_Growth_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth Index (weighted)/Year for 6 continents", y = "GDP Growth (%)")

continents_growth_df %>% ggplot(aes(x = Year, y = Avg_Growth_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth Index (avg)/Year for 6 continents", y = "GDP Growth (%)")
```




REF:

28/11/2025 - asked ChatGPT - "i have text near the last points on the x axis, how can i make it so it doesn't overlap the axis"













Plot ldc growth, each separate country in a  certain continent (6 graphs)
Create a ldc database

```{r}
ldc <- gdp_per_capita_df %>%
  filter (Ldc == TRUE)
```

Plot the graphs

```{r}
ggplot(ldc, aes(x = Year, y = Growth, color = Entity)) +
  geom_line() +
  facet_wrap(~ Continent) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme(legend.position = "none")

```

Waiiittttt on the y axis what exactly should i put? gdp, gdp per capita or growth???

7 do a weighted average of ldc growth rate for each contine, then plot a graph of each continent and how they compare

```{r}
ldc_weighted <- ldc %>%
  group_by(Year, Continent) %>%
  mutate(
    weighted_growth = sum(Gdp_per_capita * Growth, na.rm = TRUE) / sum(Gdp_per_capita, na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(ldc_weighted, aes(x = Year, y = weighted_growth, colour = Continent)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ Continent) +
  theme(legend.position = "none") +
  geom_hline(yintercept = 7, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") 


```










REF:

28/11/2025 - asked ChatGPT - "“I have a facet plot in ggplot with some continents having tons of lines and some only one. How can I adjust the y-axis scales so each graph uses the right space for its lines and they don’t look squished or empty?"

