---
title: "Introduction to Data Science Group Project Task 2"
author: "K9"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
# !!! NOTE THAT: by defalt, in the pdf file generated, it does not show the code
# chunk, if want to show chunk, change the line below to **echo = TRUE**
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(knitr)
```

In this section, we will discuss the extent each continent working toward the goal
of reducing the proportion of youth not in employment, education or training (hereafter
referred to as NEET).

We would first draw a graph to demonstrate the change of NEET for each continent 
with the year using the line graph.

```{r import data}
youth_neet <- read.csv("Exported_data/youth_neet_cleaned.csv")
```

```{r Continent NEET against Year (line chart)}
youth_neet %>%
  group_by(Continent, Year) %>%
  summarise(
    NEET_avg = weighted.mean(Percentage_NEET, Population, na.rm = T),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = Year, y = NEET_avg, colour = Continent)) +
  geom_point() +
  geom_line() +
  labs(
    x = "Year",
    y = "Percentage",
    title = "Comparison of NEET"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(
    size = 12,
    hjust = 0
  ))
```

\newpage

We then would like to see the regression line for the NEET.

```{r Continent NEET against Year (regression), message= FALSE}
youth_summary <- youth_neet %>%
  group_by(Continent, Year) %>%
  # Use na.rm = TRUE in weighted.mean to handle missing data
  summarise(
    NEET_avg = weighted.mean(Percentage_NEET, Population, na.rm = TRUE),
    .groups = "drop"
  )

# Define the color mapping vector
continent_colors <- c(
  "Africa"        = "#F8766D",
  "Asia"          = "#B79F00",
  "Europe"        = "#00BA38",
  "North America" = "#00BFC4",
  "Oceania"       = "#619CFF",
  "South America" = "#F564E3"
)
# 1. Calculate all slopes, p-values, and annotation positions
slope_data <- youth_summary %>%
  group_by(Continent) %>%
  do({
    lm_fit <- lm(NEET_avg ~ Year, data = .)
    tidy_fit <- broom::tidy(lm_fit)

    slope_stats <- tidy_fit %>% filter(term == "Year")
    slope <- slope_stats$estimate

    p_value <- slope_stats$p.value

    # Determine the significance notation based on standard alpha levels (0.05, 0.01, 0.001)
    significance_label <- case_when(
      p_value < 0.001 ~ "(p < 0.001)",
      p_value < 0.01 ~ "(p < 0.01)",
      p_value < 0.05 ~ "(p < 0.05)",
      TRUE ~ " (n.s.)" # Not significant (p >= 0.05)
    )

    # Create the final annotation label string, including the formatted slope and significance
    slope_label <- sprintf(
      "Slope: %s%.3f %s",
      ifelse(slope >= 0, "+", ""), # Add a '+' sign for positive slopes
      slope,
      significance_label
    )

    data.frame(
      Continent = unique(.$Continent),
      slope_label = slope_label
    )
  }) %>%
  ungroup()

# Section 2: Combined Plot Generation

# 2. Create the combined plot using facet_wrap
combined_plot <- ggplot(youth_summary, aes(x = Year, y = NEET_avg)) +
  geom_point(aes(colour = Continent), alpha = 0.7, size = 3) +
  geom_smooth(aes(colour = Continent, fill = Continent),
    method = "lm",
    se = TRUE, linewidth = 1
  ) +
  geom_text(
    data = slope_data,
    aes(label = slope_label, colour = Continent),
    x = -Inf, y = Inf,
    hjust = -0.1, vjust = 1.2,
    size = 3.5, fontface = "bold"
  ) +
  scale_colour_manual(values = continent_colors) +
  scale_fill_manual(values = continent_colors) +
  facet_wrap(~Continent,
    scales = "free_y"
  ) +
  labs(
    x = "Year",
    y = "Weighted Average Percentage",
    title = "Youth Not in EET: Trend by Continent"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, hjust = 0),
    legend.position = "none",
    strip.text = element_text(size = 10, face = "bold")
  )

# Display the final combined plot object
combined_plot
```

\newpage

Then we would like to see how education and employment each contribute to the NEET, 
one would create a table to illustrate how they correlated

```{r Contribution to NEET, message=FALSE}
# First, we merge the Education Attainment and youth unemployment data into the
# youth summary data frame. Ignore NAs so that the weighted average calculation
# would be more calculating from the rest of the data.
youth_summary <- youth_neet %>%
  group_by(Continent, Year) %>%
  summarise(
    EA_avg = weighted.mean(Edu_Attainment, Population, na.rm = TRUE),
    Unemp_avg = weighted.mean(Youth_Unemployment, Population, na.rm = T),
    .groups = "keep"
  ) %>%
  right_join(., youth_summary) %>%
  ungroup()


# Then we would like to calculate the correlation between NEET rate and education
# attainment rate and youth unemployment rate.

# I used AI for this part, specifically how to turn the cor.test() result into
# a data frame.

Cor_NEET_Con <- (function() {
  EA_corr <- youth_summary %>%
    group_by(Continent) %>%
    summarise(
      test = list(cor.test(NEET_avg, EA_avg, method = "spearman"))
    ) %>%
    mutate(
      Corr_EA = map_dbl(test, ~ round(.x$estimate, 4)),
      P_Value_EA = map_dbl(test, ~ round(.x$p.value, 4))
    ) %>%
    select(-test)

  Unemp_corr <- youth_summary %>%
    group_by(Continent) %>%
    summarise(
      test = list(cor.test(NEET_avg, Unemp_avg, method = "spearman"))
    ) %>%
    mutate(
      Corr_Unemp = map_dbl(test, ~ round(.x$estimate, 4)),
      P_Value_Unemp = map_dbl(test, ~ round(.x$p.value, 4))
    ) %>%
    select(-test)

  full_join(
    EA_corr, Unemp_corr,
    by = "Continent"
  )
})()

kable(Cor_NEET_Con,
  format = "markdown",
  caption = "Correlation of NEET with Youth Unemployment and Education Attainment (Continental)"
)
```

If we do not group the data by continent rather seeing them as a whole,then we 
get the following table

```{r, warning=FALSE}
Cor_NEET_World <- (function() {
  NEET_EA <- cor.test(youth_neet$Percentage_NEET,
    youth_neet$Edu_Attainment,
    method = "spearman",
    use = "complete.obs"
  )

  NEET_Unemp <- cor.test(youth_neet$Percentage_NEET,
    youth_neet$Youth_Unemployment,
    method = "spearman",
    use = "complete.obs",
  )
  
  data.frame(
    variable = c("EA", "Unemployment"),
    correlation = round(c(NEET_Unemp$estimate, NEET_EA$estimate), 4),
    p_value = round(c(NEET_Unemp$p.value, NEET_EA$p.value), 4)
  )
})()

kable(Cor_NEET_World,
  format = "markdown",
  caption = "Correlation of NEET with Youth Unemployment and Education Attainment (World)"
)
```
