---
title: "IDS Group Project Data Setting & Cleaning"
author: "K9"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(readr)
```

# Assignment brief
- Perform exploratory data analysis to assess how 6 continents in the world (all continents except Antarctica) are faring with respect to the following specific targets set out by the UN:

1. Sustain per capita economic growth in accordance with national circumstances and, in particular, at least 7 per cent gross domestic product growth per annum in the least developed countries.

2. By 2020, substantially reduce the proportion of youth not in employment, education or training.

- Find another CSV file to add to analysis

*(Please note that hereafter education, employment, and training would be referred to as EET)*

# Data Used in the Report

In the first chunk, we would import the following data:

1. The three data sets given in the Hub;

2. The additional list of the least developed countries in the world;

3. The population of each country for each year in the interval 1960 - 2024 (World Bank, 2025a);

4. The youth unemployment rate (as a percentage of the total labour force) for each country (World Bank, 2025b);

5. The education attainment rate of the population aged 25+ for at least completing an upper secondary education (World Bank, 2025c).

\* *Youth generally would be referred to the individual between age 15-24 *
```{r importing data, message=FALSE}
### Please note that if you have directly cloned the GitHub repository, then you
### can just leave the code as that, **NO need to change working directory**

### And if you would like to add additional files, please just use the path starting
### with `data_sets` (as below), as ideally the default wd is the IDS_Group_Proj already

continents <- read.csv("data_sets/continents-according-to-our-world-in-data.csv")
gdp_per_capita_df <- read.csv("data_sets/gdp-per-capita-worldbank.csv")
youth_not_in_eet <- read.csv("data_sets/youth-not-in-education-employment-training.csv")

# We now create a vector of all LDC countries
ldc_countries <- c(
  "Afghanistan", "Angola", "Bangladesh", "Benin", "Bhutan", "Burkina Faso",
  "Burundi", "Cambodia", "Central African Republic", "Chad", "Comoros",
  "Democratic Republic of the Congo", "Djibouti", "Eritrea", "Ethiopia",
  "Gambia", "Guinea", "Guinea-Bissau", "Haiti", "Kiribati",
  "Lao People's Democratic Republic", "Lesotho", "Liberia", "Madagascar",
  "Malawi", "Mali", "Mauritania", "Mozambique", "Myanmar", "Nepal",
  "Niger", "Rwanda", "Sao Tome and Principe", "Senegal", "Sierra Leone",
  "Solomon Islands", "Somalia", "South Sudan", "Sudan", "Timor-Leste",
  "Togo", "Tuvalu", "Uganda", "United Republic of Tanzania",
  "Vanuatu", "Yemen", "Zambia"
)

# Importing the population data
population <- read_csv("data_sets/population.csv", skip = 3)

# Importing the youth unemployment data
youth_unemployment <- read_csv("data_sets/youth_unemployment.csv",
  skip = 3
)


# Importing the education Attainment data
edu_attainment <- read_csv("data_sets/edu_attainment_completed_upper_secondary.csv",
  skip = 3
)
```

## Original Data Sets Given

We now clean the data, getting rid of any irrelevant data.
```{r}
# The year is irrelevant in continents, as the only thing this df needs to be
# used for is assigning a continent to a country
continents$Year <- NULL
continents$Entity <- NULL

# We now assign the continent to each country in gdp per capita
gdp_per_capita_df <- gdp_per_capita_df %>%
  left_join(continents, by = "Code")

# Same thing for youth not in eet
youth_not_in_eet <- youth_not_in_eet %>%
  left_join(continents, by = "Code")

# The continents table is now redundant (Commented this out as when running
# this code seperately without the previous chunk it would produce error)

# rm(continents)

# We clean population data

population <- population %>%
  # Remove indicator code&name, and the last column which is all NAs
  select(-`Indicator Code`, -`Indicator Name`, -70) %>%
  rename(
    Entity = `Country Name`,
    Code = `Country Code`
  ) %>%
  pivot_longer(
    cols = 3:67,
    names_to = "Year",
    values_to = "Population"
  )
```


Renaming columns to nicer names
```{r}
# Rename a column in gdp per capita
gdp_per_capita_df <- gdp_per_capita_df %>%
  rename(Gdp_per_capita = GDP.per.capita..PPP..constant.2017.international...)

# Rename a column in youth not in eet
youth_not_in_eet <- youth_not_in_eet %>%
  rename(
    Percentage_NEET =
      Share.of.youth.not.in.education..employment.or.training..total....of.youth.population.
  )
```


Dropping values before the year 2014, and if population data is above 2021, and if continent is NA.
```{r}
# Drop all values before 2014, as these UN goals were set in 2015
gdp_per_capita_df <- gdp_per_capita_df %>%
  filter(Year >= 2014 & !is.na(Continent))

youth_not_in_eet <- youth_not_in_eet %>%
  filter(Year >= 2014 & !is.na(Continent))
```



Calculating Economic Growth and change in eet of each country
```{r}
gdp_per_capita_df <- gdp_per_capita_df %>%
  mutate(Growth = ifelse(lag(Entity) != Entity, NA, (Gdp_per_capita - lag(Gdp_per_capita)) /
    lag(Gdp_per_capita) * 100))

youth_not_in_eet <- youth_not_in_eet %>%
  mutate(Perc_Change_NEET = ifelse(lag(Entity) != Entity, NA, (Percentage_NEET -
    lag(Percentage_NEET)) / lag(Percentage_NEET) * 100))
```


This chunk is now to add a flag to the gdp per capita dataframe, labelling if a country is an LDC or not
```{r}
gdp_per_capita_df <- gdp_per_capita_df %>%
  mutate(Ldc = Entity %in% ldc_countries)
```

Merging the data frames to include population in both
```{r}
gdp_per_capita_df <- merge(gdp_per_capita_df, population, by = c("Entity", "Year", "Code"))
youth_not_in_eet <- merge(youth_not_in_eet, population, by = c("Entity", "Year", "Code"))
```

Now we convert to numeric and calculate GDP per capita of each country each year
```{r}
gdp_per_capita_df$Population <- as.numeric(gdp_per_capita_df$Population)

gdp_per_capita_df <- gdp_per_capita_df %>%
  mutate(Gdp = Population * Gdp_per_capita)
```

Finally, export the CSV files of the 2 data frames
```{r}
write.csv(gdp_per_capita_df, "Exported_data/gdp_per_capita_cleaned.csv", row.names = FALSE)
write.csv(youth_not_in_eet, "Exported_data/youth_neet_cleaned.csv", row.names = FALSE)
```

## Additional Data Sets

We now clean the youth unemployment df and the education attainment df.

```{r}
youth_unemployment <- youth_unemployment %>%
  # Remove indicator code&name, and the last column which is all NAs
  select(-`Indicator Code`, -`Indicator Name`, -70) %>%
  rename(
    Entity = `Country Name`,
    Code = `Country Code`
  ) %>%
  pivot_longer(
    cols = 3:67,
    names_to = "Year",
    values_to = "Youth_Unemployment"
  ) %>%
  filter(Year >= 2014) 

youth_unemployment$Year <- as.numeric(youth_unemployment$Year)

edu_attainment <- edu_attainment %>%
  # Remove indicator code&name, and the last column which is all NAs
  select(-`Indicator Code`, -`Indicator Name`, -70) %>%
  rename(
    Entity = `Country Name`,
    Code = `Country Code`
  ) %>%
  pivot_longer(
    cols = 3:67,
    names_to = "Year",
    values_to = "Edu_Attainment"
  ) %>%
  filter(Year >= 2014) 

edu_attainment$Year <- as.numeric(edu_attainment$Year)

# Merge these two df's into the youth_not_in_eet

youth_not_in_eet <- youth_not_in_eet %>% 
  left_join(. , edu_attainment, by = c("Code", "Year", "Entity")) %>% 
  left_join(. , youth_unemployment, by  = c("Code", "Year", "Entity"))

write.csv(youth_not_in_eet, "Exported_data/youth_neet_cleaned.csv", row.names = FALSE)
```


\newpage

# Reference

World Bank (2025c). Educational attainment, at least completed upper secondary, population 25+, total (%) (cumulative) | Data. [online] World Bank Open Data. Available at: https://data.worldbank.org/indicator/SE.SEC.CUAT.UP.ZS [Accessed 1 Dec. 2025].

World Bank (2025a). Population, Total. [online] World Bank Open Data. Available at: https://data.worldbank.org/indicator/SP.POP.TOTL [Accessed 1 Dec. 2025].

World Bank (2025b). Unemployment, Youth Total (% of Total Labor Force Ages 15-24) (modeled ILO estimate) | Data. [online] World Bank Open Data. Available at: https://data.worldbank.org/indicator/SL.UEM.1524.ZS [Accessed 1 Dec. 2025].

# LLM Usage

25/11/2025 asked ChatGPT "what would the definition of "least developed countries" mean in terms of gdp per capita (PPP)"

26/11/2025 asked ChatGPT "give me a list of all countries the UN considers LDCs between 2015 and 2021, with each country in quotation marks seperated by commas"

# Limitation of the Data Sets

1. For the Youth NEET data set, there is no data for China, which may lead to a 
large error for conclusion drawn for Asia


