---
title: "Task 1"
author: "K9"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading librarys:
```{r}
#install.packages("tidyverse")
#install.packages("ggplot2")

library(tidyverse)
library(ggplot2)
```


Import data
```{r}
gdp_per_capita_df <- read.csv("Exported_data/gdp_per_capita_cleaned.csv")
gdp_per_capita_df <- gdp_per_capita_df %>%
  filter(Year >= 2015)
```


Calculate the total GDP of each continent
```{r}
continent_gdp_totals <- gdp_per_capita_df %>%
  group_by(Continent, Year) %>%
  summarise(
    Total_GDP = sum(Gdp, na.rm = TRUE)
  ) %>%
  ungroup()
```

## Target 1.1 - Sustain per capita economic growth in accordance with national circumstances

Plot a graph of GDP against time for each continent
```{r}
continent_gdp_totals %>% ggplot(aes(x = Year, y = Total_GDP, colour = Continent)) +
  geom_point() +
  geom_line() +
  labs(title = "GDP/Year for 6 continents", y = "GDP ($)")
```


I decided this graph could show more if they all started at an index of 100
```{r}
continent_gdp_totals <- continent_gdp_totals %>%
  group_by(Continent) %>%
  mutate(
    GDP_Index = Total_GDP / first(Total_GDP) * 100
  )
```



We now replot the graph, using indexed values (starting value index 100)
```{r}
continent_gdp_totals %>% ggplot(aes(x = Year, y = GDP_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_text(
  data = continent_gdp_totals %>% filter(Year == 2021),
  aes(label = round(GDP_Index, 1)),
  hjust = -0.3,
  size = 3,
  show.legend = FALSE) +
  expand_limits(x = 2022) +
  labs(title = "GDP Index/Year for 6 continents", y = "GDP Index")
```


Calculate the weighted and mean growth of each continent
```{r}
continents_growth_df <- gdp_per_capita_df %>%
  group_by(Continent, Year) %>%
  summarise(
    Weighted_Growth = sum(Growth * Gdp, na.rm = TRUE) / sum(Gdp, na.rm = TRUE),
    Avg_Growth = mean(Growth, na.rm = TRUE)
  ) %>%
  ungroup()
```


Plot the growth graphs of each continent
```{r}
continents_growth_df %>% ggplot(aes(x = Year, y = Weighted_Growth, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth (weighted)/Year for 6 continents", y = "GDP Growth (%)")

continents_growth_df %>% ggplot(aes(x = Year, y = Avg_Growth, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  labs(title = "GDP Growth (avg)/Year for 6 continents", y = "GDP Growth (%)")
```


I decided to do a similar thing with growth about index (realised is redundant, don't use in analysis)
```{r}
continents_growth_df <- continents_growth_df %>%
  group_by(Continent) %>%
  mutate(
    Avg_Growth_Index = Avg_Growth / first(Avg_Growth) * 100,
    Weighted_Growth_Index = Weighted_Growth / first(Weighted_Growth) * 100
  )
```


Plotting the graphs
```{r}
continents_growth_df %>% ggplot(aes(x = Year, y = Weighted_Growth_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 1) +
  labs(title = "GDP Growth Index (weighted)/Year for 6 continents", y = "GDP Growth (%)")

continents_growth_df %>% ggplot(aes(x = Year, y = Avg_Growth_Index, colour = Continent)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 1) +
  labs(title = "GDP Growth Index (avg)/Year for 6 continents", y = "GDP Growth (%)")
```

Now we'll plot a graph to see if top 2 GDP countries are the drivers between continental 
Making a new data frame with the top 2 countries per continent and their gdps
```{r}
top_2_per_continent <- gdp_per_capita_df %>%
  group_by(Continent, Entity) %>%
  summarise(Total_GDP = sum(Gdp, na.rm = TRUE)) %>%
  arrange(Continent, desc(Total_GDP)) %>%
  group_by(Continent) %>%
  slice(1:2) %>%
  ungroup()
```


Function which returns a dataset containing GDP growth for the top 2 countries in a continent and the continent's average growth
```{r}
get_plot_data <- function(x) {

  top_2_entities <- top_2_per_continent %>%
    filter(Continent == x) %>%
    pull(Entity)

  country_data <- gdp_per_capita_df %>%
    filter(Continent == x, Entity %in% top_2_entities)

  continent_avg <- continents_growth_df %>%
    filter(Continent == x) %>%
    rename(Growth = Avg_Growth) %>%
    mutate(Entity = "Continent Average")

  plot_data <- bind_rows(country_data, continent_avg)

  return(plot_data)
}
```

Function to plot the graphs
```{r}
plot_drivers <- function(continent) {
  data <- get_plot_data(continent)

  ggplot(data, aes(x = Year, y = Growth, colour = Entity)) +
    geom_line() +
    geom_point() +
    labs(
      title = paste("Top 2 GDP Growth vs Avg -", continent),
      y = "Growth (%)"
    )
}
```

Plotting the graphs
```{r}
plot_drivers("Africa")
plot_drivers("Asia")
plot_drivers("Europe")
plot_drivers("North America")
plot_drivers("South America")
plot_drivers("Oceania")
```



## Target 1.2 - LDC 7% Growth

Create an LDC database
```{r}
ldc <- gdp_per_capita_df %>%
  filter (Ldc == TRUE)
```


Plot GDP growth by continent
```{r}
ggplot(ldc, aes(x = Year, y = Growth, color = Entity)) +
  geom_line() +
  facet_wrap(~ Continent) +
  labs(
    title = "GDP Growth of LDCs by Continent",
    x = "Year",
    y = "GDP Growth (%)"
  )
```


Compute the weighted average GDP growth for each continent
```{r}
ldc_weighted <- ldc %>%
  group_by(Year, Continent) %>%
  summarise(
    weighted_growth =
      sum(Gdp_per_capita * Growth, na.rm = TRUE) /
      sum(Gdp_per_capita, na.rm = TRUE),
    .groups = "drop"
  )
```


Plot weighted GDP growth by continent with a 7% reference line
```{r}
ggplot(ldc_weighted, aes(x = Year, y = weighted_growth, color = Continent)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ Continent, scales = "free_y") +
  geom_hline(yintercept = 7, linetype = "dashed", color = "red") +
  labs(
    title = "Weighted GDP Growth of LDCs by Continent",
    x = "Year",
    y = "Weighted GDP Growth (%)"
  )
```


Identify which LDCs have had a growth that reached 7% and plot their growth
```{r}
ldc_7 <- ldc %>%
  group_by(Entity) %>%
  filter(any(Growth >= 7)) %>%
  ungroup()
```


Plot their graph
```{r}
ggplot (ldc_7, aes(x = Year, y = Growth, color = Entity)) + 
  geom_point() +
  geom_line() + 
  facet_wrap (~ Continent) +
  labs(
    title = "GDP Growth of LDCs that reached 7%",
    x = "Year",
    y = "GDP Growth(%)"
  )
```


Determine which countries have had the highest growth over the years
```{r}
ldc %>%
  group_by(Entity) %>%
  summarise(avg_growth = mean(Growth, na.rm = TRUE)) %>%
  arrange(desc(avg_growth)) 
```






REF:
28/11/2025 - asked ChatGPT - "i have text near the last points on the x axis, how can i make it so it doesn't overlap the axis"
28/11/2025 - asked ChatGPT - "“I have a facet plot in ggplot with some continents having tons of lines and some only one. How can I adjust the y-axis scales so each graph uses the right space for its lines and they don’t look squished or empty?"
